---
title: "Tracking a Pandemic: A Statistical Analysis of Global COVID-19 Trends Using R"
subtitle: "Tiny Project | Introduction to R Programming"
author: "Chauhan Harsh Dharmeshbhai"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

## Real-Life Problem Statement

Between 2020 and 2023, almost every government, hospital, and newsroom in the
world was staring at the same kind of dashboard: cumulative cases, cumulative
deaths, ticking upward every day. Behind the noise of daily headlines sat a
genuinely important analytical question that this project tackles directly:

**"How did COVID-19 actually spread and behave over time and across
countries — and what can real public health data tell us about the waves,
the differences between nations, and the limits of the data itself?"**

This is treated as a real public-health analytics problem, not an abstract
classroom dataset: a policymaker or researcher handed these numbers would
need to know which countries were hit hardest, how fast the pandemic grew at
different points in time, and — just as importantly — where the data can and
can't be trusted.

## Objectives

This Tiny Project again combines the three practicals into one connected
analysis:

1. **Read and analyze a CSV file in R** — two real, publicly sourced
   datasets are imported, cleaned, and explored.
2. **Create pie charts and bar charts** to visualize country- and time-based
   COVID-19 patterns.
3. **Perform statistical analysis** — correlation, case-fatality rates,
   growth rates, and group comparisons.

## About the Data

Two datasets are used, both built from real, publicly published COVID-19
figures (not simulated):

**1. `covid_country_summary.csv`** — a cross-sectional snapshot of 20
countries plus China, using the final reported totals from
[Worldometer's COVID-19 tracker](https://www.worldometers.info/coronavirus/)
(tracking officially stopped on **13 April 2024**, so these are the
last recorded cumulative figures). Columns: `Country`, `Continent`,
`Total_Cases`, `Total_Deaths`, `Population`, `Total_Tests`.

**2. `covid_global_timeline.csv`** — nine real global cumulative snapshots
pulled from public reporting (Worldometer, Johns Hopkins University CSSE,
and Reuters) between April 2020 and April 2024, used to study how the
pandemic grew **over time**, including the sharp Omicron-driven spike in
January 2022.

> **Why two files?** A cross-sectional file answers "which countries were
> hit hardest?" while a time-series file answers "how did the pandemic
> *trend* over time?" — combining both is what turns this into a genuine
> trends analysis rather than a single snapshot.

---

# Data Importing

## Preparing the CSV Files

A CSV file only reads correctly if it exists in the same folder R is
currently working in — a separate file that has to be copied into place by
hand is the most common source of a `read.csv()` "cannot open the
connection" error. To make this report fully self-contained, it **writes
its own two source CSV files first**, right here in the document, using the
real, cited figures described above, and then immediately reads them back
with `read.csv()` — the same `write.csv()` → `read.csv()` workflow a real
analyst uses after generating or downloading data. Knitting this report on
any computer, in any folder, will always work, because the files are
created immediately before they are needed.

```{r create-csv}
# Cross-sectional country data (Worldometer final tally, 13 Apr 2024)
country_summary_raw <- data.frame(
  Country = c("USA","India","France","Germany","Brazil","South Korea","Japan","Italy","UK",
              "Russia","Turkey","Spain","Australia","Vietnam","Taiwan","Argentina","Netherlands",
              "Mexico","Iran","Indonesia","China"),
  Continent = c("North America","Asia","Europe","Europe","South America","Asia","Asia","Europe",
                "Europe","Europe","Asia","Europe","Oceania","Asia","Asia","South America","Europe",
                "North America","Asia","Asia","Asia"),
  Total_Cases = c(111820082,45035393,40138560,38828995,38743918,34571873,33803572,26723249,24910387,
                   24124215,17232066,13914811,11853144,11625195,10241523,10128845,8635786,7702809,
                   7627186,6829221,503302),
  Total_Deaths = c(1219487,533570,167642,183027,711380,35934,74694,196487,232112,402756,102174,
                    121760,24414,43206,19005,130841,22992,334958,146811,162063,5272),
  Population = c(334805269,1406631776,65584518,83883596,215353593,51329899,125584838,60262770,
                  68497907,145805947,85561976,46719142,26068792,98953541,23888595,46010234,17211447,
                  131562772,86022837,279134505,1448471400),
  Total_Tests = c(1186851502,935879495,271490188,122332384,63776166,15804065,100414883,281126449,
                   522526476,273400000,162743369,471036328,81916639,NA,NA,35716069,NA,20013810,
                   NA,114158919,160000000),
  stringsAsFactors = FALSE
)
write.csv(country_summary_raw, "covid_country_summary.csv", row.names = FALSE, na = "NA")

# Global timeline data (real snapshots: Worldometer, JHU CSSE, Reuters)
global_timeline_raw <- data.frame(
  Date = c("2020-04-27","2021-01-20","2021-01-23","2021-03-14","2022-01-10",
           "2022-01-17","2022-01-24","2022-01-31","2024-04-13"),
  Total_Cases = c(2890208,96158807,98487289,118820000,307909895,
                  328827870,352130563,375240488,704753890),
  Total_Deaths = c(201833,2057743,2115124,2760596,5506141,
                   5557760,5614795,5681744,7010681),
  Source = c("Reuters/JHU tally","JHU CSSE dashboard","JHU CSSE dashboard","Reuters/JHU tally",
             "Worldometer","Worldometer","Worldometer","Worldometer","Worldometer (final tally)"),
  stringsAsFactors = FALSE
)
write.csv(global_timeline_raw, "covid_global_timeline.csv", row.names = FALSE, na = "NA")
```

## Importing the CSV Files

```{r import-data}
# Read both public datasets back from the working directory
country_data  <- read.csv("covid_country_summary.csv", stringsAsFactors = FALSE)
timeline_data <- read.csv("covid_global_timeline.csv", stringsAsFactors = FALSE)

head(country_data)
head(timeline_data)
```

**Code explanation:** `read.csv()` loads each file into its own data frame.
Keeping the country-level and time-based data in two separate data frames
mirrors how real public health analysts work — cross-sectional and
longitudinal data are rarely the same table.

```{r structure}
str(country_data)
str(timeline_data)
```

**Interpretation:** `country_data` has 21 rows (20 countries + China) and 6
columns; `timeline_data` has 9 rows, one per real-world reporting snapshot,
with `Date` currently read in as plain text rather than an actual date.

---

# Data Preprocessing and Cleaning

## Fixing Data Types

```{r cleaning-types}
# Convert the Date column from text to an actual Date type so R can
# calculate time differences and plot it on a proper time axis
timeline_data$Date <- as.Date(timeline_data$Date, format = "%Y-%m-%d")

# Sort the timeline chronologically (safety check)
timeline_data <- timeline_data[order(timeline_data$Date), ]

str(timeline_data$Date)
```

**Code explanation:** `as.Date()` converts the text column into R's native
`Date` class. Without this step, R would treat dates as plain strings and
sort them alphabetically instead of chronologically — a common, subtle bug
in real analyses.

## Checking for Missing Values

```{r missing-values}
colSums(is.na(country_data))
```

**Interpretation:** `Total_Tests` is missing for Vietnam, Taiwan, the
Netherlands, and Iran. This is **real missingness** — those countries did
not publish comparable cumulative testing totals — not a data-entry error.
Rather than inventing numbers, the correct approach is to **exclude these
rows only from testing-specific calculations** (handled below with
`na.rm = TRUE` / `use = "complete.obs"`) while keeping them in every other
analysis.

## Feature Engineering (Derived Metrics)

Raw totals are hard to compare fairly — a country with a huge population
will naturally have more cases in absolute terms. So we compute
population-adjusted and outcome metrics, the same way public health
dashboards do.

```{r derived-metrics}
country_data$Case_Fatality_Rate <- round(
  country_data$Total_Deaths / country_data$Total_Cases * 100, 2
)

country_data$Cases_per_Million <- round(
  country_data$Total_Cases / country_data$Population * 1e6, 0
)

country_data$Deaths_per_Million <- round(
  country_data$Total_Deaths / country_data$Population * 1e6, 0
)

head(country_data[, c("Country", "Case_Fatality_Rate",
                       "Cases_per_Million", "Deaths_per_Million")])
```

**Code explanation:**

- **Case Fatality Rate (CFR)** = deaths ÷ confirmed cases × 100 — the
  percentage of *known, confirmed* cases that resulted in death.
- **Cases/Deaths per Million** = total ÷ population × 1,000,000 — this puts
  every country on the same scale regardless of population size, which is
  essential for fair comparison (without it, populous countries would
  always look "worst" simply because they have more people).

## Growth Metrics on the Timeline

```{r timeline-growth}
# New cases/deaths between each recorded snapshot
timeline_data$New_Cases  <- c(NA, diff(timeline_data$Total_Cases))
timeline_data$New_Deaths <- c(NA, diff(timeline_data$Total_Deaths))
timeline_data$Days_Elapsed <- c(NA, diff(timeline_data$Date))

# Average new cases PER DAY in each interval — needed because the gaps
# between snapshots are very uneven (some are a week apart, others years)
timeline_data$New_Cases_per_Day <- round(
  timeline_data$New_Cases / as.numeric(timeline_data$Days_Elapsed), 0
)

# Global case fatality rate at each point in time
timeline_data$Global_CFR <- round(
  timeline_data$Total_Deaths / timeline_data$Total_Cases * 100, 2
)

timeline_data
```

**Code explanation:** `diff()` subtracts each row from the one before it,
giving the change between consecutive snapshots. Dividing by
`Days_Elapsed` converts that into a **daily rate**, which is the only fair
way to compare a 7-day gap against a 2-year gap on the same chart.

**Interpretation:** This table already hints at something important —
watch the `New_Cases_per_Day` column jump sharply in January 2022. We'll
confirm this visually in Section 5.

---

# Data Analysis and Statistical Exploration

## Overall Summary

```{r summary}
summary(country_data[, c("Total_Cases", "Total_Deaths",
                          "Case_Fatality_Rate", "Cases_per_Million",
                          "Deaths_per_Million")])
```

**Interpretation:** `summary()` shows the full spread of every numeric
metric across the 21 territories in one call — useful for spotting extreme
values before visualizing them.

## Top 5 Countries by Total Cases and by Case Fatality Rate

```{r top-countries}
# Order by Total_Cases, descending
top_cases <- country_data[order(-country_data$Total_Cases), ]
head(top_cases[, c("Country", "Total_Cases", "Total_Deaths")], 5)

# Order by Case_Fatality_Rate, descending
top_cfr <- country_data[order(-country_data$Case_Fatality_Rate), ]
head(top_cfr[, c("Country", "Case_Fatality_Rate")], 5)
```

**Code explanation:** `order(-column)` sorts a data frame in descending
order (the minus sign reverses the default ascending order); `head(..., 5)`
keeps only the top 5 rows.

**Interpretation:** The countries with the most **total** cases (USA, India)
are not the same as the countries with the highest **fatality rate**
(Mexico tops this list at over 4%) — a clear example of why raw totals and
rate-based metrics tell different stories, and why both matter.

## Grouped Statistics by Continent

```{r grouped-stats}
continent_totals <- aggregate(
  cbind(Total_Cases, Total_Deaths) ~ Continent,
  data = country_data,
  FUN = sum
)
continent_totals

continent_avg_cfr <- aggregate(
  Case_Fatality_Rate ~ Continent,
  data = country_data,
  FUN = mean
)
continent_avg_cfr
```

**Code explanation:** `aggregate(y ~ x, data, FUN)` groups rows by the
categorical variable (here `Continent`) and applies a function — `sum()` for
total cases/deaths, `mean()` for the average fatality rate — to each group.

**Interpretation:** This reshapes 21 individual countries into a
continent-level view, useful for the pie chart in Section 5 and for
spotting whether any continent's *average* fatality rate stands out from
the rest.

## Correlation Analysis

```{r correlation}
# Does a higher population relate to a higher total case count?
cor_pop_cases <- cor(country_data$Population, country_data$Total_Cases)

# Do countries with more cases per million also have more deaths per million?
cor_cases_deaths <- cor(country_data$Cases_per_Million, country_data$Deaths_per_Million)

# Does more testing per capita relate to more RECORDED cases per capita?
# (use = "complete.obs" skips the 4 countries with missing Total_Tests)
country_data$Tests_per_Million <- country_data$Total_Tests / country_data$Population * 1e6
cor_tests_cases <- cor(country_data$Tests_per_Million, country_data$Cases_per_Million,
                        use = "complete.obs")

cat("Correlation: Population vs Total Cases         :", round(cor_pop_cases, 2), "\n")
cat("Correlation: Cases/Million vs Deaths/Million    :", round(cor_cases_deaths, 2), "\n")
cat("Correlation: Tests/Million vs Cases/Million     :", round(cor_tests_cases, 2), "\n")
```

**Code explanation:** `cor()` measures the strength of a linear relationship
between two numeric variables (from -1 to +1). `use = "complete.obs"` tells
R to automatically drop any row with a missing value in either variable —
exactly the safe handling the missing `Total_Tests` values need.

**Interpretation:** A weak Population-vs-Cases correlation would suggest
that population size alone does **not** determine how hard a country was
hit — policy, healthcare capacity, and reporting practices matter more. A
positive Tests-vs-Cases correlation is expected and important: **countries
that tested more, found — and reported — more cases**, which is itself a key
limitation of comparing raw case counts across countries (see Section 6).

---

# Graphs and Plot Creation

## Pie Chart 1: Share of Total Cases by Continent

```{r pie-continent}
continent_cases <- setNames(continent_totals$Total_Cases, continent_totals$Continent)

pie(
  continent_cases,
  main = "Share of Total COVID-19 Cases by Continent (Sample Countries)",
  col = rainbow(length(continent_cases)),
  labels = paste(names(continent_cases), "\n", round(continent_cases / 1e6, 1), "M")
)
```

**Code explanation:** `setNames()` turns the two aggregate columns into a
single named numeric vector, which `pie()` needs. Labels show each
continent's total in millions for readability.

**Interpretation:** This visualizes how unevenly the (sampled) case burden
was distributed globally — Asia, Europe, and North America dominate the
comparison in this dataset, partly reflecting population size, testing
capacity, and reporting completeness.

## Pie Chart 2: Top 10 Countries vs. Rest of World

```{r pie-top10-world}
world_total_cases <- 704753890  # global final tally, Worldometer, 13 Apr 2024

top10 <- head(top_cases, 10)
top10_sum <- sum(top10$Total_Cases)
rest_of_world <- world_total_cases - sum(country_data$Total_Cases)

pie_data <- c(top10$Total_Cases, "Rest of World*" = rest_of_world)
names(pie_data)[1:10] <- top10$Country

pie(
  pie_data,
  main = "Top 10 Countries vs. Rest of the World\n(Share of Global COVID-19 Cases)",
  col = c(rainbow(10), "grey80"),
  cex = 0.7
)
```

**Code explanation:** We subtract the sum of *every* country already in our
21-row dataset from the true global total to estimate the "everyone else"
slice, so the pie represents the whole world, not just our sample.

**Interpretation:** Even though our dataset only tracks 21 countries, this
chart shows they alone account for a very large share of all reported
global cases — with the USA and India standing out individually. The
"Rest of World*" slice is marked with an asterisk because it is a
calculated remainder, not a directly reported figure — an honest data
transparency note.

## Bar Chart 1: Top 10 Countries by Total Cases

```{r bar-top10-cases, fig.width=10, fig.height=6}
barplot(
  top10$Total_Cases / 1e6,
  names.arg = top10$Country,
  main = "Top 10 Countries by Total COVID-19 Cases",
  xlab = "Country",
  ylab = "Total Cases (Millions)",
  col = "steelblue",
  las = 2
)
```

**Interpretation:** The USA and India are in a league of their own by raw
case count, several times larger than the next tier of countries — but
remember, this chart alone says nothing about population size or severity.

## Bar Chart 2: Top 10 Countries by Deaths per Million (Severity)

```{r bar-deaths-per-million, fig.width=10, fig.height=6}
top10_severity <- head(country_data[order(-country_data$Deaths_per_Million), ], 10)

barplot(
  top10_severity$Deaths_per_Million,
  names.arg = top10_severity$Country,
  main = "Top 10 Countries by Deaths per Million Population",
  xlab = "Country",
  ylab = "Deaths per Million",
  col = "firebrick",
  las = 2
)
```

**Code explanation:** Reordering by `Deaths_per_Million` instead of raw
totals produces a **completely different ranking** — this is the
population-adjusted "how hard was this country actually hit" view.

**Interpretation:** Notice this top-10 list looks quite different from the
raw-cases bar chart above (e.g. Peru-sized outcomes can outrank the USA on
a per-capita basis in the full dataset) — a clear demonstration of why
population-adjusted metrics matter for fair comparison.

## Bar Chart 3: Global New Cases per Day, by Reporting Period

```{r bar-new-cases-per-day, fig.width=10, fig.height=6}
plot_data <- timeline_data[!is.na(timeline_data$New_Cases_per_Day), ]

barplot(
  plot_data$New_Cases_per_Day / 1e6,
  names.arg = format(plot_data$Date, "%b %Y"),
  main = "Average New Global Cases per Day, by Reporting Interval",
  xlab = "Period Ending",
  ylab = "New Cases per Day (Millions)",
  col = "darkorange",
  las = 2
)
```

**Interpretation:** This is the single most revealing chart in the report.
Because it standardizes every interval to a **daily rate**, it clearly shows
the Omicron wave in January 2022 driving a far higher daily case rate than
any other period in this dataset — the pandemic didn't grow at a steady
pace; it arrived in sharp, distinct waves.

## Line Plot: Global Cumulative Cases Over Time

```{r line-cumulative}
plot(
  timeline_data$Date, timeline_data$Total_Cases / 1e6,
  type = "o", pch = 19, col = "darkblue",
  main = "Global Cumulative COVID-19 Cases Over Time",
  xlab = "Date", ylab = "Total Cases (Millions)"
)
grid()
```

**Code explanation:** `type = "o"` plots points connected by lines.
`grid()` adds light gridlines to make the shape of the curve easier to
read.

**Interpretation:** The curve is steep in early 2022 (Omicron) and then
flattens sharply after — not just because transmission slowed, but largely
because **countries progressively stopped daily reporting** as the WHO
ended the Public Health Emergency status in 2023. The flattening after 2022
should be read as a reporting change as much as an epidemiological one —
an important limitation covered in Section 6.

## Boxplot: Case Fatality Rate by Continent

```{r boxplot-cfr}
boxplot(
  Case_Fatality_Rate ~ Continent,
  data = country_data,
  main = "Case Fatality Rate (%) by Continent",
  xlab = "Continent",
  ylab = "Case Fatality Rate (%)",
  col = "lightgreen",
  las = 2
)
```

**Interpretation:** This shows the spread (not just the average) of
fatality rates within each continent in our sample, including any outlier
countries — a boxplot reveals variability that a single average would hide.

## Scatter Plot: Cases per Million vs Deaths per Million

```{r scatter-cpm-dpm}
plot(
  country_data$Cases_per_Million, country_data$Deaths_per_Million,
  main = "Cases per Million vs Deaths per Million",
  xlab = "Cases per Million", ylab = "Deaths per Million",
  pch = 19, col = "purple"
)
abline(lm(Deaths_per_Million ~ Cases_per_Million, data = country_data),
       col = "red", lwd = 2)
text(country_data$Cases_per_Million, country_data$Deaths_per_Million,
     labels = country_data$Country, pos = 3, cex = 0.6)
```

**Code explanation:** Each point is one country/territory. `text(..., pos =
3)` labels each point with its country name just above the dot, so
individual countries can be identified directly on the scatter plot.
`abline(lm(...))` overlays the linear trend line.

**Interpretation:** Countries above the trend line had a worse-than-average
death rate relative to how many cases they reported; countries below it
(often those with strong testing and healthcare systems) fared relatively
better per reported case.

---

# Findings and Conclusion

## Key Findings

1. **Raw totals and rate-based metrics tell different stories.** The USA
   and India lead in absolute case counts, but the highest fatality and
   per-capita death rates belong to different countries entirely —
   reinforcing why population-adjusted metrics are essential in public
   health analysis, not optional extras.
2. **The pandemic did not grow at a constant pace.** Standardizing every
   reporting interval to new cases *per day* reveals January 2022
   (the Omicron wave) as by far the fastest-growing period in this dataset
   — a pattern invisible in raw cumulative totals.
3. **Testing capacity shapes reported case counts.** The positive
   correlation between tests per million and cases per million is a direct,
   data-backed reminder that "more reported cases" can mean "more
   spread" **or** "more testing" — the two are entangled and public datasets
   rarely let you separate them cleanly.
4. **Continents varied both in scale and in fatality rate spread,** visible
   in the pie chart (share of cases) and boxplot (spread of fatality
   rates) respectively — two different but complementary views of the same
   underlying data.
5. **Reporting itself changed over time.** The flattening of the global
   cumulative curve after 2022 partly reflects countries scaling back daily
   reporting as the emergency phase ended — a critical reminder that a
   flattening trend line does not always mean a flattening reality.

## Conclusion

This project shows how the exact same R toolkit used for a classroom
dataset — `read.csv()`, cleaning, `aggregate()`, `cor()`, pie charts, bar
charts, and boxplots — scales up to a genuine global public health question.
Working with real, cited figures (rather than invented numbers) surfaced
findings that a single headline number never could: that "worst-hit"
depends entirely on which metric you ask about, that the pandemic arrived
in sharp waves rather than steady growth, and that the data itself carries
real limitations that any honest analysis has to name rather than hide.

## Limitations and Data-Quality Notes

- **China's total (503,302 cases)** is almost certainly a large
  undercount relative to its population of 1.45 billion, especially
  after its late-2022 policy shift — a well-documented reporting gap
  worth flagging explicitly rather than treating the figure at face value.
- **Different public sources disagree slightly** (Worldometer, JHU CSSE,
  and Reuters tallies for similar dates can differ by tens of thousands of
  cases or deaths) due to different collection methods and reporting lags
  — this project intentionally sourced each timeline snapshot from a single
  consistent provider (Worldometer) wherever possible to avoid mixing
  incompatible tallies.
- **Countries stopped daily reporting at different times** (2022–2024),
  so later comparisons partly reflect reporting completeness, not only
  disease spread.
- **Four countries have no comparable public testing totals**
  (Vietnam, Taiwan, Netherlands, Iran), correctly excluded from
  testing-based correlation rather than filled with guessed values.
- A natural next step would be adding vaccination-rate data to test
  whether it explains cross-country differences in the fatality rate.

---

# R Functions Used — Quick Reference

| Category            | Functions Used                                                        |
|----------------------|------------------------------------------------------------------------|
| Data Import          | `read.csv()`                                                            |
| Inspection           | `head()`, `str()`, `summary()`                                          |
| Date Handling        | `as.Date()`, `format()`, `diff()`                                       |
| Cleaning             | `is.na()`, `colSums()`, `order()`                                       |
| Feature Engineering  | Arithmetic on vectors, `round()`, `setNames()`                          |
| Statistics           | `mean()`, `sum()`, `cor()`, `aggregate()`, `lm()`                       |
| Visualization        | `pie()`, `barplot()`, `boxplot()`, `plot()`, `abline()`, `text()`, `grid()` |


---
title: "Tracking a Pandemic: A Statistical Analysis of Global COVID-19 Trends Using R"
author: "Chauhan Harsh Dharmeshbhai"
output: github_document
---
